---
title: "ethan_running_variable"
author: "Ethan Bledsoe"
date: "2026-09-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(tidyverse)
library(sf)
library(stringr)
```


## Add treatment information

```{r}
# Read treated municipalities
treated <- read_csv(
  "~/Desktop/HSIF/Causal Inference/datasets/treated61-2.csv",
  show_col_types = FALSE
) %>%
  mutate(
    CVEGEO = str_pad(
      as.character(CVEGEO),
      width = 5,
      side = "left",
      pad = "0"
    )
  )

final_key_df <- read_csv(
  "~/Desktop/HSIF/Causal Inference/datasets/final_key_df_head.csv",
  show_col_types = FALSE)

# Keep only variables needed for the join
# distinct() prevents accidental duplication of municipalities
treated_important <- treated %>%
  dplyr::select(
    CVEGEO,
    gov_inclusion_code
  ) %>%
  distinct(CVEGEO, .keep_all = TRUE)

# Add treatment information
final_key_df <- final_key_df %>%
  left_join(
    treated_important,
    by = "CVEGEO"
  ) %>%
  mutate(
    ever_treated = as.integer(!is.na(gov_inclusion_code))
  )

# Check number treated
table(final_key_df$ever_treated)

```


# Dataset being used is final_key_df

## Composite z-score: standardize, orient, average (equal and inverse-variance weights)
```{r}

# Basic z-score using scale() function
final_key_df <- final_key_df %>% 
  mutate(homicidio_scaled = as.numeric(scale(homicidio_doloso_2025))) %>% 
  mutate(robo_veh_scaled = as.numeric(scale(robo_veh_violencia_2025))) %>% 
  mutate(inseguro_scaled = as.numeric(scale(n_inseguro_popw)))

# Equal weights ranking

final_key_df <- final_key_df %>% 
  mutate(equal_score = (homicidio_scaled + robo_veh_scaled + inseguro_scaled)) %>% 
  arrange(desc(equal_score))

final_key_df <- final_key_df %>% 
  mutate(
    equal_rank = rank(-equal_score, ties.method = "min")
  )

# How many treated municipalities are in the top 61: 43
final_key_df %>%
  filter(equal_rank <= 61) %>%
  summarise(
    n_treated_in_top_53 = sum(ever_treated == 1, na.rm = TRUE)
  )

```
## Common analysis sample
```{r}
analysis_df <- final_key_df %>%
  filter(
    !is.na(homicidio_doloso_2025),
    !is.na(robo_veh_violencia_2025),
    !is.na(n_inseguro_popw)
  ) %>%
  mutate(
    homicidio_scaled = as.numeric(
      scale(homicidio_doloso_2025)
    ),
    
    robo_veh_scaled = as.numeric(
      scale(robo_veh_violencia_2025)
    ),
    
    inseguro_scaled = as.numeric(
      scale(n_inseguro_popw)
    )
  )

nrow(analysis_df)
```

## Equal-weight composite z-score

```{r}

equal_df <- analysis_df %>%
  mutate(
    equal_score = (
      homicidio_scaled +
      robo_veh_scaled +
      inseguro_scaled
    ) / 3
  ) %>%
  arrange(desc(equal_score)) %>%
  mutate(
    equal_rank = row_number()
  )

# Exact top 61
equal_top61 <- equal_df %>%
  slice_head(n = 61)

# In-sample overlap with treated municipalities
equal_top61 %>%
  st_drop_geometry() %>%
  summarise(
    n_treated_in_top_61 =
      sum(ever_treated == 1),
    
    pct_treated_in_top_61 =
      mean(ever_treated == 1) * 100
  )

```


## Logistic regression score

```{r}
logit_model <- glm(
  ever_treated ~
    homicidio_scaled +
    robo_veh_scaled +
    inseguro_scaled,
  data = analysis_df,
  family = binomial
)

summary(logit_model)

logit_df <- analysis_df %>%
  mutate(
    logit_score = predict(
      logit_model,
      type = "response"
    )
  ) %>%
  arrange(desc(logit_score)) %>%
  mutate(
    logit_rank = row_number()
  )

# Exact top 61
logit_top61 <- logit_df %>%
  slice_head(n = 61)

# In-sample overlap with treated municipalities
logit_top61 %>%
  st_drop_geometry() %>%
  summarise(
    n_treated_in_top_61 =
      sum(ever_treated == 1),
    
    pct_treated_in_top_61 =
      mean(ever_treated == 1) * 100
  )

```

## Principal component score

```{r}
pca_model <- prcomp(
  analysis_df %>%
    st_drop_geometry() %>%
    dplyr::select(
      homicidio_scaled,
      robo_veh_scaled,
      inseguro_scaled
    ),
  center = FALSE,
  scale. = FALSE
)

summary(pca_model)

# Examine PC1 loadings
pca_model$rotation[, 1]
```
```{r}
pca_score <- pca_model$x[, 1]

# Mean of the three standardized measures
overall_insecurity <- analysis_df %>%
  st_drop_geometry() %>%
  transmute(
    overall = (
      homicidio_scaled +
      robo_veh_scaled +
      inseguro_scaled
    ) / 3
  ) %>%
  pull(overall)

# Flip PC1 if necessary
if (cor(pca_score, overall_insecurity) < 0) {
  pca_score <- -pca_score
}

pca_df <- analysis_df %>%
  mutate(
    pca_score = pca_score
  ) %>%
  arrange(desc(pca_score)) %>%
  mutate(
    pca_rank = row_number()
  )

# Exact top 61
pca_top61 <- pca_df %>%
  slice_head(n = 61)

# Overlap with treated municipalities
pca_top61 %>%
  st_drop_geometry() %>%
  summarise(
    n_treated_in_top_61 =
      sum(ever_treated == 1),
    
    pct_treated_in_top_61 =
      mean(ever_treated == 1) * 100
  )
```
## Compare ranking methods
```{r}
equal_ids <- equal_top61$CVEGEO
logit_ids <- logit_top61$CVEGEO
pca_ids   <- pca_top61$CVEGEO

tibble(
  comparison = c(
    "Equal vs Logistic",
    "Equal vs PCA",
    "Logistic vs PCA"
  ),
  overlap = c(
    length(intersect(equal_ids, logit_ids)),
    length(intersect(equal_ids, pca_ids)),
    length(intersect(logit_ids, pca_ids))
  )
)
```

## Municipalities selected by logistic but not equal weights
```{r}
logit_only <- logit_top61 %>%
  st_drop_geometry() %>%
  filter(!CVEGEO %in% equal_ids) %>%
  dplyr::select(
    CVEGEO,
    NOMGEO,
    logit_rank,
    logit_score,
    ever_treated
  )

logit_only
```
## Municipalities selected by equal weights but not logistic
```{r}
equal_only <- equal_top61 %>%
  st_drop_geometry() %>%
  filter(!CVEGEO %in% logit_ids) %>%
  dplyr::select(
    CVEGEO,
    NOMGEO,
    equal_rank,
    equal_score,
    ever_treated
  )

equal_only
```

